---
title: "Quadratic Linear Mixed Model"
format: 
  html:
    toc: true
    toc-depth: 3
---

```{r}
#| include: false
library(tumr)
```



The package also includes `quad()`, which fits a quadratic linear mixed effects model to tumor growth data. This model is useful when tumor growth over time may be nonlinear rather than strictly linear.

Quadratic linear mixed models are well suited for longitudinal tumor growth data because they account for:

- Fixed effects: population-level effects of interest, such as treatment, time, and the quadratic effect of time
- Random effects: subject-specific variability that induces correlation among repeated measurements

By default, `quad()` fits the model:

$$
\log(1 + \text{measure}) \sim (\text{time} + \text{time}^2) * \text{group} + (\text{time} \mid \text{id})
$$

## Model fit
```{r}
mel2 <- tumr(melanoma2, ID, Day, Volume, Treatment)
fit_quad <- quad(mel2)
```

## Summary

```{r}
summary(fit_quad)
```


## Plot

```{r}
plot(fit_quad) + ggplot2::scale_y_log10()
plot(fit_quad, "contrast")
```